---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapfit

<!-- badges: start -->
[![R-CMD-check](https://github.com/okamumu/mapfit/workflows/R-CMD-check/badge.svg)](https://github.com/okamumu/mapfit/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/mapfit)](https://CRAN.R-project.org/package=mapfit)
[![Codecov test coverage](https://codecov.io/gh/okamumu/mapfit/branch/main/graph/badge.svg)](https://app.codecov.io/gh/okamumu/mapfit?branch=main)
<!-- badges: end -->

The goal of mapfit is to estimate parameters of phase-type distribution (PH) and Markovian arrival process (MAP). PH/MAP fitting is required in the analysis of non-Markovian models involving general distributions. By replacing general distributions with estimated PH/MAP, we have approximate the non-Markovian models with continuous-time Markov chains (CTMCs). Our tool offers

* PH/MAP fitting with grouped data
* PH fitting with theoretical probability density functions.

These features help us to analyze non-Markovian models with phase expansion.

## Installation

``` r
# Install devtools from CRAN
install.packages("mapfit")

# Or the development version from GitHub:
# install.packages("devtools")
devtools::install_github("okamumu/mapfit")
```

## PH Fitting

### Overview

PH distribution is defined as the time to absorption in a time-homogeneous CTMC with an absorbing state. The p.d.f. and cumulative distribution function (c.d.f.) are mathematically given as the expressions using matrix exponential. Let $\boldsymbol{\alpha}$ and $\boldsymbol{Q}$ denote a probability (row) vector for determining an initial state and an infinitesimal generator for transient states, respectively. Since the c.d.f. is given by the probability that the current state of underlying CTMC has already been absorbed, the c.d.f. of PH distribution is given by
$$
F(x) = 1 - \boldsymbol{\alpha} \exp(\boldsymbol{Q} x) \boldsymbol{1},
$$
where $\boldsymbol{1}$ is a column vector whose entries are 1. Also the p.d.f. can be obtained by taking the first derivative of the c.d.f.; 
$$
f(x) = \boldsymbol{\alpha} \exp(\boldsymbol{Q} x) \boldsymbol{\xi},
$$
where $\boldsymbol{\xi} = - \boldsymbol{Q} \boldsymbol{1}$.

The purpose of PH fitting is to determine PH parameters $\boldsymbol{\alpha}$ and $\boldsymbol{Q}$ so that the estimated PH distribution fits to observed data. There are two different approaches; MM (moment match) method and MLE (maximum likelihood estimation). The MM method is to find PH parameters whose first few moments match to the moments from empirical data or distribution functions. On the other hand, MLE is to find PH parameters maximizing the likelihood (probability) of which the data is drawn from the model as a sample.

### Data for PH fitting

The parameter estimation algorithm generally depends on data forms to be used. mapfit deals with several kinds of data in PH fitting; point data, weighted point data, grouped data, grouped data with missing values and truncated data. The point data consists of independent and identically-distributed (IID) random samples.

|Sample No.|Time|
|:-:|:-:|
|1 | 10.0 |
|2 | 1.4 |
|...|...|
|100 | 51.0 |


The above table shows an example of point data for a hundred IID samples. The weighted point data is the point data in which all points have their own weights. The weighted point data is used for numerical integration of a density function in our tool. The grouped data consists of break points and counts. For each of two successive break points, the number of samples is counted as a bin. This is equivalent to the data format to draw a histogram.

The grouped data with missing values allows us to use the grouped data in which several counts are unknown (missing). In the tool, missing counts are expressed by `NA`. Also, in the truncated data, several samples are truncated at a point before their observations (right censored data). The truncated data can be represented as the number of samples in a specific interval from the point to infinity in the context of grouped data.

|Time interval| Counts|
|:-:|:-:|
|[0, 10]| 1 |
|[10, 20]| `NA` |
|[20, 30]|  4 |
|[30, 40]| 10 |
|[40, 50]| `NA` |
|[50, 60]| 30 |
|[60, 70]| 10 |
|[70, 80]| 12 |
|[80, 90]| 4 |
|[90, 100]| 0 |
|[100, Inf)| 5 |

The above table shows an example of the grouped data on break points 0, 10, 20, ..., 100 where the data has missing values at the intervals [10,20] and [40,50]. Furthermore, the last 5 samples are truncated at 100.

|Time interval| Counts|
|:-:|:-:|
|[0, 10]| 1 |
|[10, 20]| `NA` |
|[20, 30]|  4 |
|[30, 40]| 10 |
|[40, 50]| `NA` |
|[50, 60]| 30 |
|[60, 70]| 10 |
|[70, 80]| 12 |
|[80, 90]| 4 |
|[90, 100]| 0 |
|[100, Inf)| `NA` |

On the other hand, the aboe table shows an example of another grouped data. In this case, several samples are truncated at 100 but we do not know the exact number of truncated samples.

### Models and Methods

PH distributions are classified to sub-classes by the structure of $\boldsymbol{\alpha}$ and $\boldsymbol{Q}$, and the parameter estimation algorithms depend on the class of PH distribution. The tool deals with the following classes of PH distribution:

- general PH: The PH distribution in which there are no constraints on $\boldsymbol{\alpha}$ and $\boldsymbol{Q}$. In the tool, this is referred to as `ph' class.
- canonical form 1 (CF1): One of the minimal representations of acyclic PH distribution. The matrix $\boldsymbol{Q}$ becomes a bidiagonal matrix whose entries are sorted in increasing order. In the tool, this is referred to as `cf1' class.
- hyper-Erlang distribution: One of the representations of acyclic PH distribution. The distribution consists of a mixture of Erlang distributions. In the tool, this is referred to as `herlang' class.

The parameters of `ph` class are $\boldsymbol{\alpha}$, $\boldsymbol{Q}$ and $\boldsymbol{\xi}$, which are defined as members of R6 class in R. To represent the matrix $\boldsymbol{Q}$, we use `Matrix` package which is an external package of R. The `cf1` class are inherited from the `ph` class. In addition to inherited members, `cf1` has a member `rate` to store the absolute values of diagonal elements of $\boldsymbol{Q}$. The `herlang` class has the member for the mixed ratio as `alpha`, shape parameters of Erlang components `shape`, rate parameters of Erlang components `rate`. `herlang` classes can be transformed to `ph` class by using `as.gph` method of R. `cf1` has members for `alpha`, `Q` and `xi`.

The R functions for PH parameter estimation are;

- `phfit.point`: MLEs for general PH, CF1 and hyper-Erlang distribution from point and weighted point data. The estimation algorithms for general PH and CF1 are the EM algorithms proposed in [1]. The algorithm for hyper-Erlang distribution is the EM algorithm with shape parameter selection described in [2,3].
- `phfit.group`: MLEs for general PH, CF1 and hyper-Erlang distribution from grouped and truncated data. The estimation algorithms for general PH and CF1 are the EM algorithms proposed in [4]. The algorithm for hyper-Erlang distribution is the EM algorithm with shape parameter selection, which is originally developed as an extension of [2,3].
- `phfit.density`: MLEs for general PH, CF1 and hyper-Erlang distribution from a density function defined on the positive domain [0, Inf). The function `phfit.density` calls `phfit.point` after making weighted point data. The weighted point data is generated by numerical quadrature. In the  tool, the numerical quadrature is performed with a double exponential (DE) formula.
- `phfit.3mom`: MM methods for acyclic PH distribution with the first three moments [5,6].

The functions `phfit.point`, `phfit.group` and `phfit.density` select an appropriate algorithm depending on the class of a given PH distribution. These functions return a list including the estimated model (`ph`, `cf1` or `herlang`), the maximum log-likelihood (llf), Akaike information criterion (aic) and other statistics of the estimation algorithm. Also, the function `phfit.3mom` returns a `ph` class whose first three moments match to the given three moments.

### Example

Here we introduce examples of the usage of PH fitting based on IID samples from Weibull distribution. At first, we load the mapfit package and generate IID samples from a Weibull distribution:
```{r wsample1}
library(mapfit)

RNGkind(kind = "Mersenne-Twister")
set.seed(1234)
wsample <- rweibull(100, shape=2, scale=1)
```
`wsample` is set to a vector including a hundred IID samples generated from Weibull distribution with scale parameter 1 and shape parameter 2. `set.seed(1234)` means determing the seed of random numbers.
```{r wsample2}
wsample
```
Based on the point data, we can estimate PH parameters. Here we obtain the estimated parameters for general PH with 5 states, CF1 with 5 states and the hyper-Erlang with 5 states by the following commands, respectively;
```{r phfit1}
## phfit with GPH
phfit.point(ph=ph(5), x=wsample)

## phfit with CF1
phfit.point(ph=cf1(5), x=wsample)

## phfit with Hyper-Erlang
phfit.point(ph=herlang(5), x=wsample, ubound=3)
```
In the above example, the number of Erlang components is restructured to 3 or less by using `ubound` argument (see [2] in detail).

Also we present PH fitting with grouped data. In this example, we make grouped data from the point data `wsample` by using the function `hist` which is originally a function to draw a histogram.
```{r gdat1}
h.res <- hist(wsample, breaks="fd", plot=FALSE)

h.res$breaks

h.res$counts
```
In the above, breaks are determined according to Freedman-Diaconis (FD) rule. Then we can get estimated PH parameters from grouped data.
```{r phfitg1}
## phfit with GPH
phfit.group(ph=ph(5), counts=h.res$counts, breaks=h.res$breaks)

## phfit with CF1
phfit.group(ph=cf1(5), counts=h.res$counts, breaks=h.res$breaks)

## phfit with Hyper-Erlang
phfit.group(ph=herlang(5), counts=h.res$counts, breaks=h.res$breaks)
```

Next we present the case where PH parameters are estimated from a density function. The density function of Weibull distribution is given by a function `dweibull`. Then we can also execute the following commands;
```{r phfit_density1}
## phfit with GPH
phfit.density(ph=ph(5), f=dweibull, shape=2, scale=1)

## phfit with CF1
phfit.density(ph=cf1(5), f=dweibull, shape=2, scale=1)

## phfit with Hyper-Erlang
phfit.density(ph=herlang(5), f=dweibull, shape=2, scale=1)
```
The last two arguments for each execution are parameters of `dweibull` function. User-defined functions are also used as density functions in similar manner.

Usually, the PH fitting with density is used for the PH expansion (PH approximation) in which known general desitributions are replaced with the PH distributions estimated from these density functions. Compared to PH fitting with samples, PH fitting with density function tends to be accurate, because density function has more information than samples. Therefore, in the case of PH fitting with density function, we can treat PH distributions with high orders without causing overfitting, i.e., it is possible to perform PH fitting even if PH has 100 states;
```{r phfit_density2}
## estimate PH parameters from the density function
(result.density <- phfit.density(ph=cf1(100), f=dweibull, shape=2, scale=1))
```
The result provides a highly-accurate approximation for Weibull distribution. However, from the viewpoint of computation time, it should be noted that only `cf1` or `herlang` with lower/upper bounds of Erlang components can be applied to PH fitting with high orders. In the above example, although the number of states is 100, the execution time is in a few seconds because of the refinement of EM algorithm [1].

If we use only point data to estimate PH paramters with high orders, the overfitting is happen.
```{r draw_plot}
## estimate PH parameters from 100 samples (overfitting example)
(result.point <- phfit.point(ph=cf1(100), x=wsample))

## 3 moments match
m1 <- gamma(1+1/2)
m2 <- gamma(1+2/2)
m3 <- gamma(1+3/2)
(result.3mom <- phfit.3mom(m1, m2, m3))

library(ggplot2)
ggplot(data.frame(x=seq(0, 3, length.out=100)), aes(x=x)) +
  stat_function(fun=dweibull, args=list(shape=2, scale=1), aes_(colour='exact')) +
  stat_function(fun=dphase, args=list(ph=result.3mom), aes_(colour='from moment matching')) +
  stat_function(fun=dphase, args=list(ph=result.point$model), aes_(colour='from point data')) +
  stat_function(fun=dphase, args=list(ph=result.density$model), aes_(colour='from density'))
```

### References

- [1] H. Okamura, T. Dohi and K.S. Trivedi, A refined EM algorithm for PH distributions, Performance Evaluation, 68(10), 938-954, 2011.
- [2] A. Thummler and P. Buchholz and M. Telek, A novel approach for phase-type fitting with the EM algorithm, IEEE Transactions on Dependable and Secure Computing, 3(3), 245-258, 2006.
- [3] A. Panchenko and A. Thummler, Efficient phase-type fitting with aggregated traffic traces, Performance Evaluation, 64, 629-645, 2007.
- [4] H. Okamura, T. Dohi and K.S. Trivedi, Improvement of EM algorithm for phase-type distributions with grouped and truncated data, Applied Stochastic Models in Business and Industry, 29(2), 141-156, 2013.
- [5] A. Bobbio and A. Horvath and M. Telek, Matching three moments with minimal acyclic phase type distributions, Stochastic Models, 21(2/3), 303-326, 2005.
- [6] T. Osogami and M. Harchol-Balter, Closed form solutions for mapping general distributions to minimal PH distributions, Performance Evaluation, 63(6), 524-552, 2006.

## MAP Fitting

### Overview

MAP (Markovian arrival process) is a stochastic point process whose arrival rates are dominated by a CTMC. The CTMC expresses the internal state of MAP called a phase process. MAP is generally defined by an initial probability vector $\boldsymbol{\alpha}$ and two matrices for infinitesimal generators $\boldsymbol{D}_0$, $\boldsymbol{D}_1$. Let $\boldsymbol{\pi}(n,t)$ be the row vector whose i-th entry is the probability that the phase process is i at time t and n arrivals occur before time t. Then we have the following differential equations:
$$
\frac{d}{dt} \boldsymbol{\pi}(n,t) = \boldsymbol{\pi}(n,t) \boldsymbol{D}_0 + \boldsymbol{\pi}(n-1,t) \boldsymbol{D}_1, \quad \text{for $n = 1, 2, \ldots$},
$$
$$
\frac{d}{dt} \boldsymbol{\pi}(0,t) = \boldsymbol{\pi}(0,t) \boldsymbol{D}_0, \quad \boldsymbol{\pi}(0,0) = \boldsymbol{\alpha},
$$
where $\boldsymbol{D}_0$ and $\boldsymbol{D}_1$ are infinitesimal generators of phase process without and with arrivals, respectively. Note that $\boldsymbol{D}_0 + \boldsymbol{D}_1$ becomes the infinitesimal generator of phase process. Similar to PH fitting, the purpose of MAP fitting is to find MAP parameters $\boldsymbol{\alpha}$, $\boldsymbol{D}_0$ and $\boldsymbol{D}_1$ so that the estimated MAP fits to observed data. In MAP fitting, there are also two approaches; MM method and MLE. The MM method for MAP is to determine MAP parameters with marginal/joint moments and k-lag correlation [1]. MLE is to find MAP parameters maximizing the log-likelihood function. We implement MLE approaches in the tool.

### Data for MAP fitting

mapfit deals with point data and grouped data in MAP fitting. The point data for MAP fitting is a time series for arrivals. The grouped data for MAP fitting consists of break points and counts which are made from a time series for arrivals.

|Arrival No.|Time (sec)|
|:-:|:-:|
|1 | 1.340 |
|2 | 1.508 |
|3 | 4.176 |
|4 | 8.140 |
|5 | 11.036|
|6 | 15.072 |
|7 | 17.892 |
|8 | 20.604 |
|9 | 22.032 |
|10 | 24.300 |
|... | ... |

The above table shows an example of point data that consists of arrival times.

|Time interval | Counts |
|:-:|:-:|
|[0, 5]| 3 |
|[5, 10]| 1 |
|[10, 15]| 1 |
|[15, 20]| 2 |
|[20, 25]| 4 |
|...|...|

The above table shows an example of grouped data. The grouped data is made from the point data by counting the number of arrivals in every 5 seconds. Note that missing values cannot be treated in MAP fitting of this version of tool.

### Models and Methods

mapfit has three classes (models) for MAP, which have different parameter estimation algorithms.

- general MAP: MAP with no constraint on parameters. This class is referred to as `map` in the tool. Also, the tool uses a Markov modulated Poisson process (MMPP) as a specific structure of `map`, which can be generated by an `mmpp` command.
- HMM (hidden Markov model) with Erlang outputs (ER-HMM): One of MAP representation where Erlang outputs correspond to time intervals between two successive arrivals [2]. In the tool, this class is referred to as `erhmm`.
- MMPP with grouped data: MMPP with approximate parameter estimation. This is referred to as `gmmpp` in the tool, and is essentially same as `mmpp` except for parameter estimation algorithm. In the parameter estimation of `gmmpp`, it is assumed that at most one phase change is allowed in one observed time interval [3].

The `map` class consists of parameters $\alpha$, $D_0$ and $D_1$, which are given by slots of S4 class in R. The `gmmpp` class also has the slots `alpha`, `D0` and `D1` as model parameters. The `erhmm` class has an initial probability vector for HMM states (`alpha`), a probability transition matrix for HMM states (`P`), the shape parameters for Erlang distribution (`shape`) and the rate parameters for Erlang distribution (`rate`). The S4 class `erhmm` can be transformed to `map` by using `as` method.

The tool has the follwoing MAP fitting functions:

- `mapfit.point`: MLEs for general MAP and ER-HMM from point data. The estimation algorithm for general MAP is the EM algorithm introduced in [4]. The algorithm for ER-HMM is the EM algorithm proposed in [2].
- `mapfit.group`: MLEs for general MAP and `gmmpp` from grouped data. Both the estimation algorithms for general MAP and `gmmpp` are presented in [3]. Note that `erhmm` cannot be used in the case of grouped data.

The functions `mapfit.point` and `mapfit.group` select an appropriate algorithm depending on the class of a given MAP. These functions return a list including the estimated model (`map`, `erhmm` or `gmmpp`), the maximum log-likelihood (llf), Akaike information criterion (aic) and other statistics of the estimation algorithm. In general, `erhmm` for point data and `gmmpp` for grouped data are much faster than general MAP.

### Example

Here we demonstrate MAP fitting with point and grouped data. The data used in this example is the traffic data; BCpAug89, which consists of time intervals for packet arrivals and is frequently used in several papers as a benchmark. We use only the first 1000 arrival times.
```{r bcpaug89}
RNGkind(kind = "Mersenne-Twister")
set.seed(1234)

data(BCpAug89)
BCpAug89
```
Using this point data, we can estimate parameters for general MAP with 5 states, MMPP with 5 states and ER-HMM with 5 states by the following commands, respectively;
```{r mapfitpint1}
## mapfit for general MAP
mapfit.point(map=map(5), x=cumsum(BCpAug89))

## mapfit for general MMPP
mapfit.point(map=mmpp(5), x=cumsum(BCpAug89))

## mapfit for ER-HMM
mapfit.point(map=erhmm(5), x=cumsum(BCpAug89))
```
In the above example, `cumsum` is a function to derive cumulative sums because `BCpAug89` provides time difference data. The estimation with `erhmm` is much faster than others.

Next we present MAP fitting with grouped data. The grouped data is made from `BCpAug89` by using `hist` function, i.e.,
```{r genmapgroup}
BCpAug89.group<-hist(cumsum(BCpAug89), breaks=seq(0,2.7,0.01), plot=FALSE)

BCpAug89.group$breaks

BCpAug89.group$counts
```
In the above, break points are set as a time point sequence from 0 to 2.7 by 0.01, which is generated by a `seq` function. Using the grouped data, we have the estimated parameters for general MAP, MMPP and MMPP with approximate estimation (`gmmpp`).
```{r mapfitall}
## mapfit for general MAP with grouped data
mapfit.group(map=map(5), counts=BCpAug89.group$counts, breaks=BCpAug89.group$breaks)

## mapfit for general MMPP with grouped data
mapfit.group(map=mmpp(5), counts=BCpAug89.group$counts, breaks=BCpAug89.group$breaks)

## mapfit for general MMPP with grouped data (approximation)
mapfit.group(map=gmmpp(5), counts=BCpAug89.group$counts, breaks=BCpAug89.group$breaks)
```

### References

- [1] G. Horvath, P. Buchholz and M. Telek, A MAP fitting approach with independent approximation of the inter-arrival time distribution and the lag correlation, Proceedings of the 2nd International Conference on the Quantitative Evaluation of Systems (QEST2005), 124-133, 2005.
- [2] H. Okamura and T. Dohi, Faster maximum likelihood estimation algorithms for Markovian arrival processes, Proceedings of 6th International Conference on Quantitative Evaluation of Systems (QEST2009), 73-82, 2009.
- [3] H. Okamura, T. Dohi and K.S. Trivedi, Markovian arrival process parameter estimation with group data, IEEE/ACM Transactions on Networking, 17(4), 1326-1339, 2009.
